---
title: "Clustering"
author: "Bartłomiej Ramotowski"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=F}
library(gridExtra)
library(factoextra)
library(flexclust)
library(fpc)
### library(clustertend)
library(hopkins)
library(cluster)
library(ClusterR)
library(stats)
library(corrplot)
```

## Introduction

Objective of this project is to gain insights into the relation between sleep patterns and lifestyle factors by using clustering algorithms. Identifying clusters of individuals with similar sleep patterns and lifestyle characteristics can contribute to the development of better strategies and recommendations for improving sleep quality and overall well-being.

Link to kaggle data set: <https://www.kaggle.com/datasets/uom190346a/sleep-health-and-lifestyle-dataset>


Dataset Columns descriptions form Kaggle:

Person ID: An identifier for each individual.

Gender: The gender of the person (Male/Female).

Age: The age of the person in years.

Occupation: The occupation or profession of the person.

Sleep Duration (hours): The number of hours the person sleeps per day.

Quality of Sleep (scale: 1-10): A subjective rating of the quality of sleep, ranging from 1 to 10.

Physical Activity Level (minutes/day): The number of minutes the person engages in physical activity daily.

Stress Level (scale: 1-10): A subjective rating of the stress level experienced by the person, ranging from 1 to 10.

BMI Category: The BMI category of the person (e.g., Underweight, Normal, Overweight).

Blood Pressure (systolic/diastolic): The blood pressure measurement of the person, indicated as systolic pressure over diastolic pressure.

Heart Rate (bpm): The resting heart rate of the person in beats per minute.
Daily Steps: The number of steps the person takes per day.

Sleep Disorder: The presence or absence of a sleep disorder in the person (None, Insomnia, Sleep Apnea).

### Reading Data

In this project at first we have to prepare and clean the data for clustering. We begin with reading the data from a CSV file, examining its structure, and summarizing key information. We also convert certain variables to the appropriate data types.

```{r}
health<-read.csv("Sleep_health_and_lifestyle_dataset.csv")
head(health,10)
```
```{r}
#chcking data structure
str(health)

#data summary
summary(health)
```

```{r}
#checking data for null values
health<-health[complete.cases(health), ]
```

```{r}
#converting chr to factor
health$BMI.Category<-as.factor(health$BMI.Category)
health$Blood.Pressure <-as.factor(health$Blood.Pressure)

#converting factor to integer
health$BMI.Category<-as.integer(health$BMI.Category)
health$Blood.Pressure <-as.integer(health$Blood.Pressure)
```

```{r}
health_org<-health
health<-health[c(3,5,6,7,8,9,10,11,12)]

```

## Exploratory data anlysis

Now we will examine the dataset, identify anomalies, and gain insights through visualizations. This process helps us understand the data's structure, select relevant features, and prepare for subsequent clustering algorithms.

### Variables plots

```{r}

for (i in colnames(health)) {
  hist(health[,i], main = paste("Plot of ", i))
}
```

Data plots looks good, we cant see anything unusual.

### Corelation matrix

```{r}
health_matrix <- data.matrix(health, rownames.force = NA)
M <- cor(health_matrix)
corrplot(M, method = "number", number.cex = 0.75, order="hclust")
```

In the dataset, we observe some positively correlated variables. There is a high correlation between the quality of sleep and sleep duration, as well as between daily steps and physical activity level. Additionally, we can identify some negatively correlated variables, such as the quality of sleep and stress level, or stress level and sleep duration.

We won't remove any of these variables because they contain valuable information. Deleting them from further analysis solely based on correlation levels might not be advisable.


### Scaling values

To effectively use clustering algorithms we need to scale our data. This process ensures that all variables contribute equally to the clustering analysis.

```{r}
#scaling values
health_scaled<-scale(health)
```

### Assessing Clustering Tendency

```{r}

get_clust_tendency(health_scaled, 2, graph=T)

```

Based on hpkins stat value which is above 0.95, we can assume that significant clusters exist.

## Clustering


### Optimal number of clusters

In this section focusing on clustering, we aim to identify the optimal number of clusters within the health dataset. We will evaluate the kmeans, pam,  hcutwith silhouette method to determine the most suitable cluster configuration.


```{r,echo=F}

##clustering

a <- fviz_nbclust(health_scaled, FUNcluster = kmeans, method = "silhouette") + theme_classic() 
b <- fviz_nbclust(health_scaled, FUNcluster = cluster::pam, method = "silhouette") + theme_classic() 
c <- fviz_nbclust(health_scaled, FUNcluster = hcut, method = "silhouette") + theme_classic() 
# e <- fviz_nbclust(health_scaled, FUNcluster = cluster::fanny, method = "silhouette") + theme_classic() 
```

```{r}
grid.arrange(a, b, c, ncol=2)
```

In further analysis we will focus on clustering with k-means using 4 or 6 clusters, PAM with 6 clusters and hierarchical clustering with 7 clusters.


### Using clustering algorithms

#### K-means

At firs we will try clustering with k-means spliting data into 4 or 6 clusters. It seems to have fair silhouette value and we

```{r}
health_clusters_kmeans_4<- kmeans(health_scaled, 4)
health_clusters_kmeans_6<- kmeans(health_scaled, 6)
```

```{r}

fviz_cluster(list(data=health_scaled, cluster=health_clusters_kmeans_4$cluster), 
             ellipse.type="norm", geom="point", stand=FALSE, palette="jco", ggtheme=theme_classic())

```
```{r}

fviz_cluster(list(data=health_scaled, cluster=health_clusters_kmeans_6$cluster), 
             ellipse.type="norm", geom="point", stand=FALSE, palette="jco", ggtheme=theme_classic())

```

```{r}
sil<-silhouette(health_clusters_kmeans_4$cluster, dist(health_scaled))
fviz_silhouette(sil)
```

```{r}
sil<-silhouette(health_clusters_kmeans_6$cluster, dist(health_scaled))
fviz_silhouette(sil)
```

##### General characteristics of clusters
```{r}
health_clusters_kmeans_4$size
health_clusters_kmeans_4$cluster
health$cluster_kmeans_4<-health_clusters_kmeans_4$cluster

health_clusters_kmeans_6$size
health_clusters_kmeans_6$cluster
health$cluster_kmeans_6<-health_clusters_kmeans_6$cluster

```


```{r}

colnames(health)
for (i in colnames(subset(health, select = -c(cluster_kmeans_4,cluster_kmeans_6)))) {
  print(i)
  print(aggregate(health, health[[i]] ~ cluster_kmeans_4, mean))
}

```




#### PAM

```{r}
health_clusters_pam_6<- pam(health_scaled, 6)
```

```{r}

fviz_cluster(health_clusters_pam_6, 
             data = health_scaled, 
             elipse.type = 'convex', 
             main='Pam 6',
             show_labels = FALSE)

```

```{r}
fviz_silhouette(health_clusters_pam_6,ncol=2)
```
##### General characteristics of clusters
```{r}
health_clusters_pam_6$size
health_clusters_pam_6$cluster
health$cluster_pam_6<-health_clusters_pam_6$cluster

```


```{r}

colnames(health)
for (i in colnames(subset(health, select = -c(cluster_kmeans_4,cluster_kmeans_6,cluster_pam_6)))) {
  print(i)
  print(aggregate(health, health[[i]] ~ cluster_pam_6, mean))
}

```

#### hierarchical

```{r}
health_clusters_diana <- eclust(health_scaled, k=6, FUNcluster = 'diana', hc_method = 'ward.D')

pltree(health_clusters_diana, hang = -1, main = "DIANA", cex = 0.47)
rect.hclust(health_clusters_diana, k=6, border='red')

health_clusters_diana_stats <- cluster.stats(dist(health_scaled), health_clusters_diana$cluster)
health_clusters_diana_stats$avg.silwidth
health_clusters_diana_stats$cluster.size
```


After testing k-means, PAM, and hierarchical clustering, the highest silhouette score among our attempts was achieved by PAM with 6 clusters.

```{r}

colnames(health)
for (i in colnames(subset(health, select = -c(cluster_kmeans_4,cluster_kmeans_6,cluster_pam_6)))) {
  print(i)
  print(aggregate(health, health[[i]] ~ cluster_pam_6, mean))
}

```


```{r}
for (i in c(1:6) ){
  print(i)
  print(summary(health[health$cluster_pam_6 == i,1:9]))
}
```

Different clusters reveal variations in sleep patterns, physical activity levels, stress , daily steps, and BMI categories. Some individual has shorter but better quality sleep, with high physical activity levels and a constant stress level. Others have longer and more consistent sleep, maintaining lower stress levels and low physical activity levels.

## Conclusions 

We successfully clustered health, sleep and lifestyle dataset using the PAM algorithm and validated it with the silhouette index. We gained valuable insights into the data, revealing notable differences between the clusters. 

 



