---
title: "Association rules"
author: "Bartłomiej Ramotowski"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
library(arules)
library(arulesViz)
```

# Introduction

The goal of this project is to determine relations between sleep patterns, lifestyle choices and health metrics. Association rules helps us understand the complex connections in our dataset. We want to uncover valuable information that can guide people in sleeping better, making healthier choices, and feeling their best.

Link to kaggle data set: <https://www.kaggle.com/datasets/uom190346a/sleep-health-and-lifestyle-dataset>


Dataset Columns descriptions form Kaggle:

Dataset Columns descriptions form Kaggle:

Person ID: An identifier for each individual.

Gender: The gender of the person (Male/Female).

Age: The age of the person in years.

Occupation: The occupation or profession of the person.

Sleep Duration (hours): The number of hours the person sleeps per day.

Quality of Sleep (scale: 1-10): A subjective rating of the quality of sleep, ranging from 1 to 10.

Physical Activity Level (minutes/day): The number of minutes the person engages in physical activity daily.

Stress Level (scale: 1-10): A subjective rating of the stress level experienced by the person, ranging from 1 to 10.

BMI Category: The BMI category of the person (e.g., Underweight, Normal, Overweight).

Blood Pressure (systolic/diastolic): The blood pressure measurement of the person, indicated as systolic pressure over diastolic pressure.

Heart Rate (bpm): The resting heart rate of the person in beats per minute.
Daily Steps: The number of steps the person takes per day.

Sleep Disorder: The presence or absence of a sleep disorder in the person (None, Insomnia, Sleep Apnea).

## Reading data

```{r read}
health<-read.csv("Sleep_health_and_lifestyle_dataset.csv")
str(health)
```

```{r summary}

summary(health)

```

```{r}

health<-health[complete.cases(health), ]

```

### Columns

In this project, the prescription of right categories was crucial, as it establishes a strong foundation for use of the association rules alghorithms.

Variables categories are based on variable descriptions from Kaggle. I have also attached information about websites that I relied on while creating these categories.

#### Age
```{r}

summary(health$Age)
health$Age =  ifelse(health$Age<18, "Age below 18", 
              ifelse(health$Age<=24, "Age 18-24",
              ifelse(health$Age<=34, "Age 25-34",
              ifelse(health$Age<=44, "Age 35-44",
              ifelse(health$Age<=54, "Age 45-54",
              ifelse(health$Age<=64, "Age 55-64",
              ifelse(health$Age>=65, "Age 65 or above", NA)))))))

```

#### Sleep time

Sleep time in our data has min=5.8 and max=8.5 recommended sleep time for adults is 7-9 hourshttps://www.nhlbi.nih.gov/health/sleep/how-much-sleep 
In our data there are only adults.

```{r}

summary(health$Sleep.Duration)

health$Sleep.Duration =  ifelse(health$Sleep.Duration<7, "Too short sleep", 
                         ifelse(health$Sleep.Duration<=9, "Enought sleep",
                         ifelse(health$Sleep.Duration>9, "Too long sleep"))) 
        
```

#### Sleep Quality

```{r}
summary(health$Quality.of.Sleep)

health$Quality.of.Sleep =  ifelse(health$Quality.of.Sleep<=4, "Low quality sleep", 
                           ifelse(health$Quality.of.Sleep<=6, "Medium quality sleep",
                           ifelse(health$Quality.of.Sleep<=8, "Good quality sleep",
                           ifelse(health$Quality.of.Sleep>8, "Very good quality sleep",NA))))
```

#### Physical activity level

```{r}
summary(health$Physical.Activity.Level)

health$Physical.Activity.Level =  ifelse(health$Physical.Activity.Level<=30, "Under 30 min of physical activity a day", 
                                  ifelse(health$Physical.Activity.Level<=60, "30-60 min of physical activity a day",
                                  ifelse(health$Physical.Activity.Level>60, "Over 60 min of physical activity a day",NA)))
```

#### Stress level
```{r}
summary(health$Stress.Level)

health$Stress.Level =  ifelse(health$Stress.Level<=2, "Very low stressed",
                       ifelse(health$Stress.Level<=4, "Low stressed", 
                       ifelse(health$Stress.Level<=6, "Medium stressed",
                       ifelse(health$Stress.Level<=8, "Stressed",
                       ifelse(health$Stress.Level>8, "Very stressed",NA)))))
```

#### Blood Preasue

```{r}
summary(health$Blood.Pressure)
```

#### Heart Rate (bpm):

The resting heart rate of the person in beats per minute. <https://www.physio-pedia.com/Heart_Rate>

```{r}
summary(health$Heart.Rate)

health$Heart.Rate =  ifelse(health$Heart.Rate<=55, "Excellent heart rate", 
                     ifelse(health$Heart.Rate<=65, "Good heart rate",
                     ifelse(health$Heart.Rate<=75, "Avrage heart rate",
                     ifelse(health$Heart.Rate<=85, "Poor heart rate",       
                     ifelse(health$Heart.Rate>85, "Bad heart rate",NA)))))
                   
```

#### Daily Steps

<https://www.10000steps.org.au/articles/healthy-lifestyles/counting-steps/>

```{r}
summary(health$Daily.Steps)

health$Daily.Steps =  ifelse(health$Daily.Steps<=5000, "Sedentary daily steps", 
                      ifelse(health$Daily.Steps<=7499, "Low daily steps",
                      ifelse(health$Daily.Steps<=9999, "Somewhat active daily steps",
                      ifelse(health$Daily.Steps<=12499, "Active daily steps",       
                      ifelse(health$Daily.Steps>12500, "Highly active daily steps",NA)))))
```

#### BMI Category 
```{r}
summary(health$BMI.Category)
unique(health$BMI.Category)
health$BMI.Category = ifelse(health$BMI.Category=="Normal", "Normal weight",
                      ifelse(health$BMI.Category=="Overweight", "Overweight",
                      ifelse(health$BMI.Category=="Obese", "Obese",NA)))
```
#### Sleep disorder 
```{r}
summary(health$Sleep.Disorder)
unique(health$Sleep.Disorder)         
health$Sleep.Disorder= ifelse(health$Sleep.Disorder=="None", "No sleep disorder",
                      ifelse(health$Sleep.Disorder=="Sleep Apnea", "Sleep Apnea",
                      ifelse(health$Sleep.Disorder=="Insomnia", "Insomnia",NA)))
```

### Saving the data 
```{r}
health_clean=health[c(2:13)]
str(health_clean)
write.csv(health_clean, file = 'health_clean.csv')

```

## Association rules

```{r}
transactions = read.transactions('health_clean.csv', format='basket', sep=',', skip=0)
inspect(transactions[1:10])
size(transactions) 
length(transactions)
```



```{r}
round(itemFrequency(transactions),3) %>% sort(.,decreasing=T) %>% head(10)
```

### Standard rules

```{r}

rules.transactions<-apriori(transactions, parameter=list(supp=0.1, conf=0.5))
rules.transactionsSorted<-sort(rules.transactions, by="lift", decreasing=TRUE)
inspect(head(rules.transactionsSorted))

rules_head <- head(rules.transactionsSorted,10)
plot(rules_head, method="graph", measure="support", shading="lift")
```


Now, let's examine some basic rules to gain insights into individuals who are "stressed," "overweight," and have "insomnia."

### Stressed rule

#### Right hand side (consequent)

```{r}
rulesStressed<-apriori(data=transactions, parameter=list(supp=0.1, conf=0.1), appearance=list(default="lhs", rhs="Stressed"), control=list(verbose=F)) 
rulesStressedSorted<-sort(rulesStressed, by="confidence", decreasing=TRUE)
inspect(head(rulesStressedSorted))

rules_head <- head(rulesStressedSorted,10)
plot(rules_head, method="graph", measure="support", shading="lift")
```

People with sleep apnea or too short sleep are likely to be stressed. Similar associations exist for high blood pressure, low daily steps, overweight, and too short sleep. 

#### Left hand side (antecedent)

```{r}
rulesStressed<-apriori(data=transactions, parameter=list(supp=0.1, conf=0.1), appearance=list(default="rhs", lhs="Stressed"), control=list(verbose=F)) 
rulesStressedSorted<-sort(rulesStressed, by="confidence", decreasing=TRUE)
inspect(head(rulesStressedSorted))

rules_head <- head(rulesStressedSorted,10)
plot(rules_head, method="graph", measure="support", shading="lift")
```

These rules indicate relationships between stress and to short sleep, medium sleep quality,average heart rate, being overweight and being male. 

### Overweight rule

#### Right hand side (consequent)

```{r}
rulesOverweight<-apriori(data=transactions, parameter=list(supp=0.1, conf=0.1), appearance=list(default="lhs", rhs="Overweight"), control=list(verbose=F)) 
rulesOverweightSorted<-sort(rulesOverweight, by="confidence", decreasing=TRUE)
inspect(head(rulesOverweightSorted))

rules_head <- head(rulesOverweightSorted,10)
plot(rules_head, method="graph", measure="support", shading="lift")
```

Haveing blood pressure of 140/95, being a nurse, having sleep apnea, engaging in over 60 minutes of daily physical activity, being female, or having an average heart rate are all linked to being overweight.

#### Left hand side (antecedent)

```{r}
rulesOverweight<-apriori(data=transactions, parameter=list(supp=0.1, conf=0.1), appearance=list(default="rhs", lhs="Overweight"), control=list(verbose=F)) 
rulesOverweightSorted<-sort(rulesOverweight, by="confidence", decreasing=TRUE)
inspect(head(rulesOverweightSorted))

rules_head <- head(rulesOverweightSorted,10)
plot(rules_head, method="graph", measure="support", shading="lift")
```

These rules suggest connections between being overweight and too short sleep, average heart rate, low daily steps, being female, and the absence of sleep disorders

### Insomnia rule

#### Right hand side (consequent)

```{r}
rulesInsomnia<-apriori(data=transactions, parameter=list(supp=0.1, conf=0.1), appearance=list(default="lhs", rhs="Insomnia"), control=list(verbose=F)) 
rulesInsomniaSorted<-sort(rulesInsomnia, by="confidence", decreasing=TRUE)
inspect(head(rulesInsomniaSorted))

rules_head <- head(rulesInsomniaSorted,10)
plot(rules_head, method="graph", measure="support", shading="lift")
```

Basing on results being 35-44 years old,low daily steps, overweight, and having too short sleep gives likelihood (89.6%) of  experience insomnia.

#### Left hand side (antecedent)

```{r}
rulesInsomnia<-apriori(data=transactions, parameter=list(supp=0.1, conf=0.1), appearance=list(default="rhs", lhs="Insomnia"), control=list(verbose=F)) 
rulesInsomniaSorted<-sort(rulesInsomnia, by="confidence", decreasing=TRUE)
inspect(head(rulesInsomniaSorted))

rules_head <- head(rulesInsomniaSorted,10)
plot(rules_head, method="graph", measure="support", shading="lift")
```

These rules highlight connections between insomnia and too short sleep (which is not suprising), 30-60 min of physical activity a day, being overweight, in age between 35-44, and average heart rate.

## Conclusions 

We successfully applied association rules to identify relationships among characteristics of various health variables. Furthermore, we analyzed individuals experiencing stress, being overweight, and having insomnia. This approach provided valuable insights into the health data.
